---
title: "Exploratory Data Analysis in R with the `tidyverse`"
subtitle: "Day 1 Walkthrough"
author: "Dr. Chester Ismay"
format: 
  html:
    toc: true
    toc-title: "Contents"  # Optional, customize the title of the TOC
    toc-location: left     # Specifies the location of the TOC (can also be 'right')
    toc-float: true        # Enables the floating TOC
    toc-depth: 3           # Sets the depth of the TOC (levels of headers)
---

```{r}
#| include: false
options(width = 120)
```


# Day 1: Exploring, Cleaning, and Organizing Data

## Session 1: Foundations of EDA in R with the tidyverse

### 1: Installing the necessary packages 

```{r}
#| eval: false

# Create a vector of package names needed for this course
packages <- c(
  "moderndive",          # For tidyverse-based data analysis tools and datasets
  "fivethirtyeight",     # Contains curated datasets from FiveThirtyEight.com
  "knitr",               # For formatting outputs and rendering R Markdown documents
  "janitor",             # For cleaning data and column names easily
  "plotly",              # For creating interactive visualizations
  "dplyr",               # For data wrangling: filter(), mutate(), summarize(), etc.
  "ggplot2",             # For static data visualizations using the grammar of graphics
  "tidyr",               # For tidying and reshaping data
  "stringr",             # For working with strings using consistent functions
  "forcats",             # For working with factors (especially useful with categorical variables)
  "lubridate",           # For working with date and time data
  "purrr")               # For functional programming tools to iterate and map over data

# Install all the packages listed above from CRAN using a reliable mirror
install.packages(packages, repos = "https://cran.rstudio.com")

# Alternative: You could install a bundled version of these packages using the tidyverse meta-package
# packages_revised <- c(
#   "moderndive", "fivethirtyeight", "knitr", "janitor", "plotly", 
#   "tidyverse")  # This includes ggplot2, dplyr, tidyr, stringr, forcats, purrr, etc.

# install.packages(packages_revised, repos = "https://cran.rstudio.com")

# Install the azflights24 package from GitHub (not available on CRAN)
install.packages("remotes")  # Make sure the remotes package is available
remotes::install_github("moderndive/azflights24")  # Install from GitHub directly
```


### 2. Loading the tidyverse and data

```{r}
# Load packages we will need


# I'll try to be explicit with using :: as we get started though so
# you know which package includes the function used.

# Load the spotify_by_genre dataset from the moderndive package

```


### 3. Viewing the data

```{r}
# View the structure of the dataset including variable types and example values

```

```{r}
#| eval: false
# Open the dataset in a spreadsheet-style viewer (interactive, only works in RStudio)

```


### 4. Remove songs listed multiple times in the data

```{r}
# Remove duplicate songs based on the track_name column
# .keep_all = TRUE keeps the first full row for each unique track_name


# View the resulting cleaned dataset

```

### 5. Take a random sample of the data

```{r}
# Set a random seed so results are reproducible


# Take a random sample of 10 rows from the full dataset

```

```{r}
# Take a random sample of 10 rows but focus only on selected columns

```


### 6. Spot-checking unique genres

```{r}
# Count the number of tracks in each genre and sort from most to least

```


### 7. Checking for missing values

```{r}
# For each column, count how many missing (NA) values there are

```

### 8. Identifying numeric vs categorical variables

```{r}
# Check the class of the entire dataset (it's a tibble/data frame)

# Check the class of a specific column: track_genre (should be character or factor)

# Check the class of a numeric column: popularity (should be numeric or integer)

# Use purrr::map_dfr() to apply class() to every column and return results as a data frame

# Glimpse the result for a readable, wide-format view of column types

```


### 9. Creating summary tables

```{r}
# Use dplyr to compute common summary statistics for the popularity column
# (mean, median, and standard deviation)


# Use moderndive's tidy_summary() to get a full summary of one or more columns

```


### 10. Quick plot of distribution

```{r}
# Create a histogram to visualize the distribution of track popularity scores

```


---

### Session 1 Review Questions

**(1.1)** Why is Exploratory Data Analysis (EDA) considered a crucial first step in any data science project?

A. It allows you to train models more quickly.  
B. It helps reveal patterns, identify data issues, and guide analysis decisions.  
C. It replaces the need for data cleaning by automating it.  
D. It transforms all variables into numeric format for modeling.

---

**(1.2)** What is the role of the `distinct()` function in preparing your data?

A. It filters out rows with missing values.  
B. It removes duplicate rows based on selected columns.  
C. It transforms all character variables into factors.  
D. It visualizes the distribution of numeric variables.

---

**(1.3)** What does the `spotify_by_genre` dataset contain?

A. Only audio features from 6,000 classical music tracks  
B. Survey results about Spotify users’ listening habits  
C. Metadata and audio features of 6,000 songs across six genres  
D. Real-time streaming counts of trending tracks globally

---

**(1.4)** Which function is best used to take a random subset of rows from a dataset?

A. `sample_n()`  
B. `slice_sample()`  
C. `randomize()`  
D. `filter()`  

---

**(1.5)** What is the purpose of using `purrr::map_dfr()` in this session?

A. To count the number of missing values in a column  
B. To classify songs into popular and non-popular categories  
C. To apply a function across all columns and return the results in a single data frame  
D. To create visualizations for each genre using ggplot2

---


## Session 2: Data Cleaning Fundamentals with `dplyr`, `tidyr`, and `janitor`

### 11. Filtering and selecting columns

```{r}
# Select a subset of columns and rename track_genre to genre
# Then take a random sample of 20 rows to explore

```

```{r}
# Filter for only songs with popularity greater than 80
# Select only key identifying columns, randomly sample 20, then 
# arrange by popularity (descending)

```


### 12. Creating conditional flags

```{r}
# Create a new logical column to flag songs with both high energy and 
# high danceability

```

```{r}
# Create a new categorical column for popularity levels using case_when
# Then count how many songs fall into each category

```


### 13. Creating new columns with `mutate()`

```{r}
# Create a new column that calculates the ratio of energy to danceability

```

```{r}
# Add a logical column that flags songs with tempo greater than 140 as high tempo
# Then count how many songs fall into each category

```


### 14. Grouping and summarizing

```{r}
# Group the data by genre and calculate the average energy for each group

```

```{r}
# Count how many songs are popular or not within each genre

```


### 15. Wide to long with `tidyr::pivot_longer()`

```{r}
# Reshape selected audio features from wide format to long format
# This creates a 'feature' column for the variable name and a 'value' 
# column for its value

```


### 16. Long to wide with `tidyr::pivot_wider()`

```{r}
# Count the number of songs in each genre by popularity category
# Then reshape the data to have separate columns for 'popular' and 'not popular'

```


### 17. Splitting a column into multiple parts with `tidyr::separate()`

```{r}
# Find the maximum number of artists in any single song by counting semicolons
# Add 1 since semicolons separate the names (n artists = n semicolons + 1)

```

```{r}
# Identify the song that has the most artists

```

```{r}
# Create a vector of new column names based on the maximum number of artists found

```

```{r}
# Use separate() to split the 'artists' column into multiple artist columns
# Use fill = "right" to handle cases with fewer than max_artists
# Keep the original 'artists' column for reference

```

```{r}
# Summarize how many multi-artist songs exist by genre

```


### 18. Uniting columns

```{r}
# Combine the track_name and artists columns into a single column called track_artist
# Use " by " as the separator between the song title and the artist(s)

```


### 19. Cleaning up with `janitor`

```{r}
#| eval: false
# Simulate messy column names by renaming clean ones to include emojis, symbols, and inconsistent formatting
spotify_unruly <- spotify |>
  dplyr::rename(
    `Track Name 🎵` = track_name,
    `ARTISTS (main + featured)` = artists,
    `Album-Name___v2` = album_name,
    `DURATION (ms)` = duration_ms,
    `Popularity SCORE (%)` = popularity,
    `track genre!` = track_genre,
    `Explicit?` = explicit,
    `IS_IT_POPULAR????` = popular_or_not
  ) |> 
  dplyr::mutate(valence = NA_real_)  # Add an entirely empty column to test janitor functions

# Check the column names and dimensions before cleaning

```

```{r}
# Clean the messy column names: make them lowercase, snake_case, and remove special characters
# Remove any completely empty columns


# Check the new cleaned column names and dimensions

```


### 20. Reordering and recoding variables

```{r}
# Move the track_genre column to appear right before track_id
# Recode popular_or_not into a new column popular_recoded using case_match
# Place the new column directly after track_genre

```

---

### Session 2 Review Questions

**(2.1)** What is the purpose of using `filter()` in data wrangling with `dplyr`?

A. To remove missing values from all columns.  
B. To select specific columns by name.  
C. To remove duplicate rows from a dataset.  
D. To return only rows that meet a certain condition.

---

**(2.2)** What is the result of using the `mutate()` function in a `dplyr` pipeline?

A. It permanently deletes rows that contain NA values.  
B. It creates new columns or modifies existing ones.  
C. It filters the dataset by logical conditions.  
D. It combines multiple columns into one.

---

**(2.3)** What does the following code do in the context of the `spotify_by_genre` dataset?

```r
spotify |>
  mutate(high_acoustic_mellow = acousticness > 0.8 & valence < 0.4)
```

A. It filters the dataset to only include mellow and acoustic songs.  
B. It creates a new variable indicating whether a song is both highly acoustic and low in valence.  
C. It summarizes acousticness and valence for mellow songs.  
D. It creates a histogram of acousticness for songs with low valence.

---

**(2.4)** What is the effect of `pivot_longer()` on a dataset?

A. It removes NA values from multiple columns.  
B. It splits a character column into multiple columns.  
C. It summarizes grouped values into wider format tables.  
D. It transforms columns into key-value pairs, increasing the number of rows.

---

**(2.5)** Why is `janitor::clean_names()` helpful in the data cleaning process?

A. It changes all character variables to numeric format.  
B. It fills in missing values in your dataset.  
C. It converts messy column names to consistent, lowercase, snake_case format.  
D. It rearranges the rows of the dataset based on alphabetical order.

---

## Session 3: Managing Dates, Strings, and Categories

### 21. Working with strings using `stringr`

```{r}
# Create a new column that counts the number of characters in each track name
# Then display the track names sorted by length in descending order

```

```{r}
# Detect songs where the title includes the word "love" (case-insensitive)
# Use str_to_lower() to ensure consistent matching

```

```{r}
# Count how many "love" songs appear in each genre
# Calculate the percentage of total songs that mention "love" for each genre

```

### 22. Extracting and replacing substrings

```{r}
# Remove all parentheses from track names

```



### 23. Wrangling artists into long format

```{r}
# Use separate_rows() to split multiple artists into individual rows
# This turns collaborations into separate entries for each artist

```

```{r}
# Count how many times each artist appears across all tracks
# Artists involved in collaborations will be counted multiple times

```


### 24. Cleaning and recoding categories

```{r}
# Use fct_lump() to group all but the top 3 most common genres into "Other"

```

```{r}
# View the current counts for popular_or_not to check the ordering and distribution

```

```{r}
# Reorder factor levels for popular_or_not so that "popular" appears first

```


### 25. Counting words in track names

```{r}

# Create a new column that counts the number of words in each track name

  
  # Group by genre and calculate the average word count per genre

  
  # Reorder genres based on average word count for clearer plotting

  # Create a horizontal bar chart of average word counts by genre

  
  # Add the rounded average word count at the end of each bar

  # Flip coordinates so genres appear on the y-axis

  
  # Add labels and title

  
  # Apply a clean minimal theme

```



### 26. Checking for artists appearing in multiple genres

```{r}
# Split artists so each appears in their own row (for collaborations)

  
  # Keep only distinct artist–genre pairs
  
  # Group by artist and filter for those appearing in more than one genre

  
  # Sort alphabetically and display artist–genre combinations

```


### 27. Parsing dates with `lubridate` and `azflights24`

```{r}
# Load the azflights24 and lubridate packages for flight data and date manipulation


# Preview a random sample of flight records

```

```{r}
# Create a proper date column by combining year, month, and day

```

```{r}
# Extract the month name (abbreviated) from the flight_date column for PHX-origin flights

```


### 28. Reordering factors for better visualizations with `forcats` and `ggplot2`

#### Example 1

```{r}
# Filter for flights that originated in PHX
# Count how many flights went to each destination
# Reorder destinations so the bars display in order of frequency

```

```{r}
# Filter for flights from PHX again
# Group rare destinations with fewer than 1000 flights into an "Other" category
# Reorder factor levels by count and visualize as a bar chart

```


#### Example 2

```{r}
# Reorder track genres by the median popularity of their songs
# Create a violin plot with quantile lines to show distribution

```


### 29. Interactive line plot of 6-hour wind speed averages

```{r}
# Prepare weather data for PHX and FLG airports during July 2024
# Bin timestamps into 6-hour intervals, create human-readable labels

```

```{r}
# Create a ggplot line chart with interactive tooltips using plotly


# Render the plot as an interactive Plotly chart with custom tooltips

```



### 30. Summarizing weather app preferences by region

```{r}
# Load survey data on weather app preferences from the fivethirtyeight package


# Prepare the data by filtering out missing values
# Lump less common weather sources into "Other"
# Reorder regions based on frequency of responses

```

```{r}
# Create a side-by-side bar plot showing weather source preferences by region

```


---

### Session 3 Review Questions

**`(3.1)`** Why does the regex pattern `"\\(.*?\\)"` used in `str_remove_all()` work better than `"\\(.*\\)"` when cleaning track names?

A. The `.*?` forces R to skip anything in parentheses rather than match it.  
B. `.*?` is a greedy matcher that ensures the longest match possible is removed.  
C. `"\\(.*?\\)"` only works when the parentheses are numeric, like `(1)` or `(2)`.
D. `.*?` makes the match non-greedy, stopping at the first closing parenthesis.  

---

**`(3.2)`** What does this code return?

```r
spotify |>
  separate_rows(artists, sep = ";") |>
  distinct(artists, track_genre) |>
  group_by(artists) |>
  filter(n_distinct(track_genre) > 1)
```

A. It returns artists who appear in more than one track.  
B. It returns tracks that are associated with more than one genre.  
C. It returns genres that feature more than one artist.  
D. It returns artists that appear in more than one genre.

---

**`(3.3)`** What will the following code produce?

```r
spotify |>
  mutate(track_name_clean = str_remove_all(track_name, "\\(.*?\\)")) |>
  select(track_name, track_name_clean)
```

A. A data frame with only tracks that contain parentheses in their names  
B. A new column with all track names replaced by text inside parentheses  
C. A new column where all content inside parentheses has been removed from track names  
D. A subset of songs where all punctuation has been stripped from the track name

---

**`(3.4)`** Which of the following best explains the use of `fct_lump(track_genre, n = 3)`?

A. It removes the three most frequent levels in the `track_genre` factor.  
B. It converts the `track_genre` variable into numeric rank based on popularity.  
C. It keeps only the top three most common genres and lumps the rest into "Other".  
D. It randomly selects three genres to collapse into one category.

---

**`(3.5)`** What does this `ggplot()` show?

```r
spotify |>
  mutate(track_genre = fct_reorder(track_genre, popularity, .fun = max)) |>
  ggplot(aes(x = track_genre, y = popularity)) +
  geom_violin()
```

A. A violin plot sorted alphabetically by genre name  
B. A violin plot of genre popularity, with genres ordered by max popularity  
C. A violin plot of max track length across genres  
D. A histogram showing popularity grouped by genre

